---
title: "Homework Week 12"
author: "Annanina Koster"
date: "5 januari 2016"
output: html_document
---


---


The homework [7 | 2016]

The data sleepstudy (lme4) allow to measure the effect of sleep deprivation of reaction time.

The average reaction time per day for subjects in a sleep deprivation study. On day 0 the subjects had their normal amount of sleep. Starting that night they were restricted to 3 hours of sleep per night. The observations represent the average reaction time on a series of tests given each day to each subject.

With the use of the sleepstudy:

- Plot the relation between Reaction and Days for each subject independently.
- Build a model with random coefficients for linear trend between Reaction and Days.
- Extract individual slopes for the relation from the model with random effects. Present summary statistics   of these slopes (1Q, median, 3Q).


---


First, the data should be loaded and inspected. There are 180 observations. All persons have been observed 10 times, one time every day for 10 days (0-9). Their mean reaction was 298.5.

```{r, warning=FALSE, message=FALSE}

library(lme4)
library(lattice)
summary(sleepstudy)
dat <- sleepstudy
table(dat$Subject)

```

---

For each Subject, the relation between Reaction and Days looks as follows:

```{r, warning=FALSE, message=FALSE}

xyplot(Reaction ~ Days | Subject, dat, type = c("g","p","r"), index = function(x,y) coef(lm(y ~ x))[1])

```

It seems that the relation between Reaction and Days is positive for all Subjects. This could be expected, since it makes sense that the less one sleeps, the slower one will respond. However, this relationship is different for all subjects. This should be included in the model. Moreover, it can be seen that for most subjects the relation in pretty linear! Hence, a linear trend can be modeled!

---

I believe that the most suitable model for this dataset would have the variable Days as a fixed effect, because we want to know the effect of the number of days without sleep. Moreover, we would like to distinguish the data per subject. As can be seen in the plots below, log(Reaction) does not make a lot of difference for the normality of the model. It also shows from the ANOVA test.

```{r}

par(mfrow=c(1,2))
hist(dat$Reaction)
hist(log(dat$Reaction))

mdl0 <- lmer(log(Reaction) ~ Days + (1|Subject), data=dat)

mdl1 <- lmer(Reaction ~ 1 + (1|Subject), data=dat) #model with slope only
summary(mdl1)
#here we can see a very high intercept and large differences between subjects
#moreover, we want to know the linear effect of days, so I include it in the second model

mdl2 <- lmer(Reaction ~ Days + (1|Subject), data=dat)
summary(mdl2)
#here we see a much lower intercept, fewer differences between subjects and a slope of 10 for days

anova(mdl0, mdl2)
#it can be seen that the use of log(Reaction) does not improve the model significantly, hence I will not

anova(mdl1, mdl2)
#it is shown that the second model is significantly better than the first (for a low p-value always choose the model with the highest degrees of freedom)

mdl3 <- lmer(Reaction ~ Days + (Days|Subject), data=dat)
summary(mdl3)
#in this model we also see that the slope of Days may differ per Subject
#this model is expected to be even better

anova(mdl2, mdl3)
#model 3 is preferred over model 2!

mdl4 <- lmer(Reaction ~ Days + (Days-1|Subject) + (1|Subject), data=dat)
summary(mdl4)

anova(mdl3, mdl4)
#it is found that model 3 is the most preferred model!

```

The 'best' model is the model I was expecting!

---

In order to extract the slopes of this relation per Subject:

```{r}

fixef(mdl3)
head(ranef(mdl3)$Subject)
randomCoefs = t(t(ranef(mdl3)$Subject) + fixef(mdl3))
randomCoefs
#but we need just the slopes, so:
slopes <- ranef(mdl3)$Subject[,2] + fixef(mdl3)[2]
slopes
summary(slopes)
boxplot(slopes) #to visualize

```

---

In conclusion, almost all subjects got a slower reaction as they got less sleep, which is what we expected since it makes sense!