---
title: "02_lab"
output: html_document
---

```{r}
install.packages('archivist')
library(archivist)
library(agricolae)
library(ggplot2)
```



```{r}
spines <- archivist::aread("pbiecek/ModeleLiniowe/repo/17df9d1fc4f43962a91b9ab14b7b8e6f")
head(spines)
levels(spines[,1])
```

```{r}
boxplot(length~Group, data=spines)
tapply(spines$length, spines$Group, var)
```

### One-Way ANOVA

```{r}
model.matrix(lm(length~Group, data=spines))
```

```{r}
apply(model.matrix(lm(length~Group, data=spines)), 2, sum)
```

```{r}
mouse.lm <-lm(length~Group, data=spines) 
mouse.anova <- anova(mouse.lm)
mouse.anova
```

```{r}
mouse.aov <- aov(length~Group, data=spines)
summary(mouse.aov)
```

Różnice w grupach są istotne statystycznie.

### Post-Hoc
Tukey test:

```{r}
TukeyHSD(mouse.aov)
HSD.test(mouse.aov, "Group", console=T)
plot(TukeyHSD(mouse.aov))
```

Significant differences: TG- vs TG+, WT+ vs TG+, WT- vs TG+. Two groups: {TG+} and {TG-, WT+, WT-}.

```{r}
LSD.test(mouse.aov, "Group", console=T)
```

Basically also two groups, but {TG-, WT+, WT-} is not homogeneous - there might be a small effect of either being transgenic or subject to treatment. 

```{r}
SNK.test(mouse.aov, "Group", console=T)
```

Two groups as in HSD test.

### Test Power

```{r}
group.means <- c(0.01, 0.0, 0.0)
group.sizes <- c(100, 100, 100)
group.sd <- 0.1
p.vals <- numeric(1000)
for(r in 1:1000){
group.data <- lapply(1:3, function(i) rnorm(n=group.sizes[i], mean=group.means[i], sd=group.sd))
group.data <- unlist(group.data)
group.data <- data.frame('obs'=group.data, 'group'=rep(1:3, group.sizes))
group.anova <- anova(lm(obs~group, group.data))
group.pval <- group.anova$Pr[1]
p.vals[r] <- group.pval
}
test.power <- sum(p.vals <= 0.05)/1000
qplot(p.vals, geom="histogram", main=test.power)
```

If there is one different group, then the test power drops when this group is smaller than the others. This is very important e.g. in medical treatment or in toxycological analyses: if the control group is larger than the test group, then the results may be unreliable and the healing (or toxic) effect may not be discovered. For example, for group centers `0.01, 0.0, 0.0` and group sizes `100, 10000, 10000` the power is `0.067`, and for the same centers but group sizes `100, 100, 100` the power is `0.103`.

An interesting conclusion: Limiting the size of the control group increases the power of the test!


