---
title: "MichalFrej_Homework5"
author: "Micha³ Frej"
date: "Thursday, November 05, 2015"
output: html_document
---
I assumed, that we have the same number of observations for all 6 groups(M.pre,M.1,M.2,F.pre,F.1,F.2).
```{r}
gender <- c("F","M")
diab <- c("pre","1","2")
delta1 <- 1
delta2 <- c(0,0,0.5) #for men
delta3 <- c(0,0,1) # for female
sigma <- 1
alpha <- 0.05
```
Underneath we have definition of function named "simulation",which calculates the power for null hypothesis H0: delta2=0 and delta3=0. Parameter n is size of particular group(f.e. M.1) and K is number of repeated simulations.

```{r}
simulation <- function(n=33,K=5000){
      df <- data.frame(sex = sort(rep(gender,times=length(diab)*n)),
                       diabetes = rep(diab,times=length(gender)*n),
                       diab_nr = rep(1:length(diab),times=length(gender)*n),
                       blood_pressure=0)
      
      powers <- sapply(1:K,function(i){
            df[df$sex=="F",]$blood_pressure <- rnorm(length(diab)*n,
                                                    mean = delta3[df[df$sex=="F",]$diab_nr],sd=sigma)
            df[df$sex=="M",]$blood_pressure <- rnorm(length(diab)*n,
                                                    mean = delta2[df[df$sex=="M",]$diab_nr],sd=sigma)
            model <- aov(blood_pressure~sex/diabetes, data=df)
            summary(model)[[1]][2,5]
      })
      
      H0_power <- function(level=alpha){
            1-sum(powers>=level)/K
      }
      
      H0_power()
}

```
I checked that for first few n function "simulation" increases. Let's see the results
```{r}
set.seed(8)
first_n <- 2
smallest_n <- sapply(first_n:18,function(i){simulation(i,4000)})
smallest_n
result <- min(which(smallest_n>=0.8))+first_n-1
```
To conclude, our smallest dataset contains 
```{r}
result * length(diab) * length(gender)
```
patients.