---
title: "Praca domowa 12"
author: "Agnieszka Sitko"
output:
  html_document:
    toc: yes
  pdf_document:
    toc: yes
---


```{r, warning = FALSE, message = FALSE, echo = FALSE}
library(lme4)
library(nlme)
library(lattice)
library(PBImisc)
library(ggplot2)
library(ade4)

spdata <- read.table("http://www.ats.ucla.edu/stat/r/faq/thick.csv", header = T, sep = ",")
attach(spdata)
```

##First model 

First, I will create a really simple model, that may be presented by a formula: `thick ~ soil`. Statistic calculated for such a model are gathered below.

```{r, warning = FALSE, message = FALSE, echo = FALSE}
fixedModel <- lme(fixed = thick ~ soil ,data = spdata, random = ~ 1 | dummy)
summary(fixedModel)
```

P-value calculated for `soil` shows that the relation between this variable and `thick` is significant. Estimated parameter shows that this relation is positive. 

Dependence between `thick` and `soil` may be seen in a plot.

```{r, warning = FALSE, message = FALSE, echo = FALSE}
ggplot(spdata, aes(x = soil, y = thick)) + 
      geom_point(shape = 1) +   
      geom_smooth(method = lm, se = FALSE) +
      ggtitle("Soil - thick scatterplot")
```

##Second model

Now, let's see whether there is any significant spatial correlation between `thick` values.

```{r, warning = FALSE, message = FALSE, echo = FALSE}
distSpatial <- dist(spdata[,2:3])
distThick <- dist(spdata[,4])

mantel.rtest(distSpatial, distThick)
```

Mantel test confirms the existance of spatial correlation. Let's built a model based on this assumption. To do so, I will include five different spatial correlation structres. Likelihoods for each model are presented below.

```{r, warning = FALSE, message = FALSE, echo = FALSE}
modelExp <- update(fixedModel, correlation = corExp(1, form = ~ east + north), method = "ML")
modelGaus <- update(fixedModel, correlation = corGaus(1, form = ~ east + north), method = "ML")
modelSpher <- update(fixedModel, correlation = corSpher(1, form = ~ east + north), method = "ML")
modelLin <- update(fixedModel, correlation = corLin(1, form = ~ east + north), method = "ML")
modelRatio <- update(fixedModel, correlation = corRatio(1, form = ~ east + north), method = "ML")

logLik <- c(logLik(fixedModel),logLik(modelExp),logLik(modelGaus),
            logLik(modelSpher),logLik(modelLin),logLik(modelRatio))
names(logLik) <- c("-", "corExp", "corGaus", "corSphere", "corLin", "corRatio")
logLik
```

So it looks like the best fit is obtained with a rational quadratic spatial correlation structure. Let's analyze the result of this model.

```{r, warning = FALSE, message = FALSE, echo = FALSE}
summary(modelRatio)
```

We may observe that a model with a spatial correlation structure rejects significance of `soil` variable.

##Conclusion

Simple linear model with just one independent variable claims that there is a noticeable relation between `thick` and `soil`. However, models with exponential, gaussian, linear and rational quadratic correlation structure denies this relation. All of those model provide better fit for out data (according to the likelihood value). Therefore, we may say, that in our case it is crucial to include spatial correlation to get more reliable conclusions.

