---
title: "Homework Week 13"
author: "Annanina Koster"
date: "12 januari 2016"
output: html_document
---

---

The homework [14 I 2016]

Let's download a toy dataset with two interesting variables, namely thick and soil. We would like to check if there is any relation between these two variables.

Try two models: simple linear regression model with just one dependent variable and a model which takes into account spatial structure of measurements (variables east and north).

Compare these two models and answer the question: is there a significant relation between variables thick (y) and soil (dependent variable)?

See this page for some hints.

---

Sources:

http://www.ats.ucla.edu/stat/r/faq/spatial_regression.htm
http://www.ats.ucla.edu/stat/r/faq/variogram_lme.htm
https://rawgit.com/pbiecek/LinearModels/master/MIMUW_2015/MaterialyPomocnicze/13_spatial.html


- - -


For this homework, I will be looking at a dataset with two interesting variables: thick and soil. The question to answer is whether there is a significant relationship between these two (with thick being the dependent variable and soil the independent). In order to answer this question, two models will be compared: a simple linear regression model and a model which takes into account the spatial structure of the measurements (using the variables east and north).

---

First of all, I will load and inspect the dataset. Then, I will plot the relation between the important variables thick and soil. It seems that there isn't a very obvious relation between the two. Maybe a slightly positive one. Afterwards, I will fit the simplest linear model as model0. As can be seen in the summary, the relation between thick and soil is significantly different from 0 at a 5% confidence level. Another simple linear model is given by model1, which uses the dummy variable that is 1 for all oberservations as the random effect. (As expected there is almost no random effect) As can also be seen from this summary, the relation would now also be significant at a 5% confidence level.

```{r, warning=FALSE, message=FALSE}
spdata <- read.table("http://www.ats.ucla.edu/stat/r/faq/thick.csv", header = T, sep = ",")
head(spdata) #this dataset already has a dummy variable that equals one for all observations
plot(spdata$soil, spdata$thick)

library(nlme)

model0 <- lm(thick~soil, data=spdata)
summary(model0)

model1 <- lme(thick ~ soil, data=spdata, random= ~1|dummy, method="ML")
summary(model1)

```

---

Since these models don't really seem to agree with the plot, I will now try to include the other two variables: east and north, to account for the spatial structure of measurements. Different types of spatial correlation structures will be tried, and the best model will be selected.

```{r}

model2 <- update(model1, correlation = corExp(form = ~ east + north), method = "ML")
summary(model2) #AIC: 169.1041, BIC: 180.6915, logLik: -79.55203
model3 <- update(model1, correlation = corGaus(form = ~ east + north), method = "ML")
summary(model3) #AIC: 91.50733, BIC: 103.0948, logLik: -40.75366
model4 <- update(model1, correlation = corSpher(form = ~ east + north), method = "ML")
summary(model4) #AIC: 344.3183, BIC: 355.9058, logLik: -167.1592
model5 <- update(model1, correlation = corLin(form = ~ east + north), method = "ML")
summary(model5) #AIC: 167.4697, BIC: 179.0571, logLik: -78.73483
model6 <- update(model1, correlation = corRatio(form = ~ east + north), method = "ML")
summary(model6) #AIC: 56.02011, BIC: 67.60755, logLik: -23.01006

```

As can be seen from these summaries, the model with rational quadratics as the spatial correlation structure is best. Hence, this will be the final model. In this model, the p-value of soil is quite high. Therefore it can be concluded that the model that takes into account the spatial structure of the measurements (rational quadratics) is the best model for this dataset, and it shows that there is no significant relationship between the important variables thick and soil.
