---
title: "Homework for 14 I 2016"
author: "Micha³ Frej"
output:
      html_document:
      toc: yes
pdf_document:
      toc: yes
---
Let's take a look at our data. First six rows are
```{r,message=FALSE,warning=FALSE}
library("nlme")
library("ade4")
spdata <- read.table("http://www.ats.ucla.edu/stat/r/faq/thick.csv", header = T, sep = ",")
head(spdata)
```

Graphically it looks like there is no significant dependancy between thick and soil.
```{r,message=FALSE,warning=FALSE}
plot(thick~soil, data=spdata,col="blue")

modelSimple <- lm(thick~soil, data=spdata)
a=as.numeric(modelSimple$coefficients[2])
b=as.numeric(modelSimple$coefficients[1])
curve(a*x+b,min(spdata$soil),max(spdata$soil),add=TRUE,col="red",lwd=2)
```

However simple linear model tells us, that slope is nonzero.
```{r,message=FALSE,warning=FALSE}
summary(modelSimple)
```

Assumptions of linear model are fulfilled, as residuals are normal.
```{r,message=FALSE,warning=FALSE}
shapiro.test(residuals(modelSimple))
qqnorm(residuals(modelSimple))
qqline(residuals(modelSimple),col="orange")
```

Our intuition tells us, that the dependant variable should depend on geographical distance. By looking at variogram and Mantel test we'll check if our guess is good.
```{r,message=FALSE,warning=FALSE}
mantel.rtest(dist(spdata[,2:3]),dist(spdata$thick),nrepet=9999)
```

Simulated p-value is very low, so probably there is a relation between distance and thick. 
```{r,message=FALSE,warning=FALSE}
plot(Variogram(spdata$thick,dist(spdata[,2:3])),xlab="Euclidean distance",span=0.4)
```

After a close look at this plot we see that for distances <=60 there is a linear trend. Therefore it's reasonable to add geographical distance to our model. We'll do this by setting different correlation matrix.
```{r,message=FALSE,warning=FALSE}
modelGaus <- lme(fixed = thick ~ soil, data = spdata, random = ~ 1 | dummy,correlation = corGaus(1, form = ~ east + north), method = "ML")

modelSpher <- lme(fixed = thick ~ soil, data = spdata, random = ~ 1 | dummy,correlation = corSpher(1, form = ~ east + north), method = "ML")

modelExp <- lme(fixed = thick ~ soil, data = spdata, random = ~ 1 | dummy,correlation = corExp(1, form = ~ east + north), method = "ML")

modelLin <- lme(fixed = thick ~ soil, data = spdata, random = ~ 1 | dummy,correlation = corLin(1, form = ~ east + north), method = "ML")

modelRatio <- lme(fixed = thick ~ soil, data = spdata, random = ~ 1 | dummy,correlation = corRatio(1, form = ~ east + north), method = "ML")

stats <- function(model){
      c(AIC(model),BIC(model),logLik(model))
}

comp <- data.frame(Simple=stats(modelSimple),Gaus=stats(modelGaus),
                  Spher=stats(modelSpher),Exp=stats(modelExp),
                  Lin=stats(modelLin),Ratio=stats(modelRatio))
rownames(comp) <- c("AIC","BIC","LogLikelihood")
comp <- t(comp)
comp
```

The best AIC,BIC and log likelihood has modelRatio. Let's look at details of this model to see, that it has much better estimates of intercept and slope then simple linear model(it has smaller standard deviation of this coefficients).
```{r,message=FALSE,warning=FALSE}
summary(modelRatio)
```

This model(p-value equal to 0.8094) tells us, that the ralation between thick and soil is not significant.
Now we'll check whether modelRation is a proper mixed model.
```{r,message=FALSE,warning=FALSE}
shapiro.test(residuals(modelRatio))
qqnorm(residuals(modelRatio))
qqline(residuals(modelRatio),col="purple")
```

The qqplot looks badly, but the sample size is quite small. Shapiro-Wilk test does not reject the null hypothesis(residuals are normal) on the level 5%.

The modelRadio model is better for our data than simple linear model. In model with spatial structure the soil variable is not significant, therefore I think, that there is no important relation between soil and thick.