---
title: "Homework 12"
author: "Katarzyna Kanska"
date: "14.01.2016"
output: 
  html_document:
    toc: TRUE
---

## Goal

Let's download a toy dataset with two interesting variables, namely `thick` (thickness of coal) and `soil` (soil quality). We would like to check if there is any relation between these two variables.

Try two models: simple linear regression model with just one dependent variable and a model which takes into account spatial structure of measurements (variables `east` and `north`).

Compare these two models and answer the question: is there a significant relation between variables `thick` (y) and `soil` (dependent variable)?

## Model without considering spatial structure

Firstly, we load the dataset.

```{r, warning=FALSE, message=FALSE}
spdata <- read.table("http://www.ats.ucla.edu/stat/r/faq/thick.csv", header = T, sep = ",")
head(spdata)
```

Then we estimate the simplest possible model:

```{r, warning=FALSE, message=FALSE}
model0 <- lm(thick ~ soil, data = spdata)
summary(model0)
```

We see that the variable is barely significant and the R^2^ is very small.

## Model considering spatial structure

In this section we will check whether taking into account the spacial structure of measurements would improve the fit.

Let us introduce the spacial structure via changing the covariance matrix for the random noise. We will consider several types of correlation and compare them.

```{r, warning=FALSE, message=FALSE}
library(nlme)
model1 <- lme(fixed = thick ~ soil, data = spdata, random = ~ 1 | dummy, method = "ML")
model1.exp <- update(model1, correlation = corExp(form = ~ east + north), method = "ML")
model1.gaus <- update(model1, correlation = corGaus(form = ~ east + north), method = "ML")
# model1.lin <- update(model1, correlation = corLin(form = ~ east + north), method = "ML")
# Error: false convergence (8)
model1.ratio <- update(model1, correlation = corRatio(form = ~ east + north), method = "ML")
model1.sphere <- update(model1, correlation = corSpher(form = ~ east + north), method = "ML")

Criteria <- function(model){
  return( data.frame(AIC = AIC(model), BIC = BIC(model), logLik = logLik(model)) )
}

models <- list( Exponential = model1.exp,
                Gaussian = model1.gaus,
#                Linear = model1.lin,
                Ratio = model1.ratio,
                Spherical = model1.sphere
                )

criteria <- t(matrix(ncol=4,unlist(lapply(models, Criteria))))
rownames(criteria) <- c("Exponential", "Gaussian", "Ratio", "Spherical")
colnames(criteria) <- c("AIC", "BIC", "logLik")
criteria[order(criteria[,3], decreasing = TRUE),]
```

All three criteria concludes that inverse quadratic correlation structure fits best.

```{r, warning=FALSE, message=FALSE}
summary(model1.ratio)
```

Here we see that the `soil` variable is definitely not significant. So including information about spatial structure makes a big difference (for simple linear model we may conclude that `soil` is significant). Therefore it is crucial to test the significance of spatial autocorrelation.

```{r, warning=FALSE, message=FALSE}
coal.dists <- dist(cbind(spdata$east, spdata$north))
thick.dists <- dist(spdata$thick)
library(ade4)
mantel.rtest(thick.dists, coal.dists, nrepet = 9999)

coal.dists <- as.matrix(coal.dists)
coal.dists.inv <- 1/coal.dists
diag(coal.dists.inv) <- 0
library(ape)
Moran.I(spdata$thick, coal.dists.inv)$p.value
```

In both Mantel and Morgan tests we can reject the null hypothesis that there is zero spatial autocorrelation. So the spacial structure is significant and should be included in the model.

## Conclusions

* In the linear model with no spacial structure included we can say that for significance level equal to `0.05` the relation between `thick` and `soil` is significant. But on the other hand only 8% of `thick` variance is explained by variation of `soil`, so the fit is very poor.
* The correlation between `thick` and place of measurement turned out to be significant, therefore we need to include this information into the model. While taken into consideration, it changes the significance of `soil` variable.
* **There is no significant relation between `thick` and `soil`**.
* This data provides a good example how excluding significant information completely changes the conclusions from the model (significant vs insignificant relation between variables).