---
title: "PracaDom12_MBogdańska"
author: "Magdalena Bogdańska"
date: "Modele liniowe i mieszane"
output: 
  html_document:
    toc: TRUE
---

## Task

Verify whether there is any relation between variables "thick" and "soil" in dataset "toy".

Try two models: simple linear regression model with just one dependent variable and a model which takes into account spatial structure of measurements (variables east and north).

Compare these two models and answer the question: is there a significant relation between variables thick (y) and soil (dependent variable)?

-------------------------------------------------------------------------------------------
```{r, warning=FALSE, message=FALSE,echo=FALSE}
library(PBImisc)
library(lattice)
library(ggplot2)
library(geoR)
library(nlme)
spdata <- read.table("http://www.ats.ucla.edu/stat/r/faq/thick.csv", header = T, sep = ",")
```

Data to be considered is of a following form:
```{r, warning=FALSE, message=FALSE}
head(spdata)
```
It contains the thickness and qualities of coal seams measured at different coordinates. 

## Model1
What is the relation between thickness and quality of soil without taking into account any spatial variable? Let us see it graphically.

```{r, message=FALSE, warning=FALSE, fig.width=8, fig.height=5}
ggplot(spdata, aes(soil,thick)) +
  geom_point() +
  stat_smooth(method="lm")+
  geom_smooth(se=FALSE, color="red", size=2,method="lm")
```

I verify first linear regression model: thick~soil.
```{r, warning=FALSE, message=FALSE}
model1 <- lm(thick~ soil, data=spdata)
summary(model1)
```
Variable soil is significant in linear regression model.

However the relation between two variables considered doesn't seem to be purely linear.
```{r, message=FALSE, warning=FALSE, fig.width=8, fig.height=5}
ggplot(spdata, aes(soil,thick)) +
  geom_point() +
  stat_smooth(method="loess")+
  geom_smooth(se=FALSE, color="red", size=2,method="loess")
```

## Model2

Let us see graphically whether there is any dependence of thickness on the geographical location. I will plot the variogram.
```{r, warning=FALSE, message=FALSE, fig.width=8, fig.height=5}
v1 <- variog(coords = spdata[,2:3], data = spdata$thick, message=FALSE)
plot(v1, type = "b", main = "Variogram for thickness")
```

From the variogram one can clearly see that there is relation between thickness of soil and distance. Thus in the subsequent model I will take into account the spatial structure of measurements (variables east and north, describing latitude and longitude).
There are several different forms that the spatial autocorrelation can take, here I will consider exponential, Gaussian, spherical and linear forms. I will compare model without spatial components with model with different spatial autocorrelation structures.

```{r, warning=FALSE, message=FALSE}
model0 <- lme(thick ~ 1, data = spdata, random = ~ 1 | dummy)

modelExp <- update(model0, correlation = corExp(1, form = ~ east + north), method = "ML")

modelGaus<- update(model0, correlation = corGaus(1, form = ~ east + north), method = "ML")

modelSpher <- update(model0, correlation = corSpher(1, form = ~ east + north), method = "ML")

modelLin <- update(model0, correlation = corLin(1, form = ~ east + north), method = "ML")

c(AIC(model0), AIC(modelExp), AIC(modelGaus), AIC(modelSpher), AIC(modelLin))
```
Based on the Akaike inforation criterion I choose Gaussian correlation structure.


I will now compare model with and without spatial structure considering also variable soil as fixed variable.
```{r, warning=FALSE, message=FALSE}
modelGausFinal <- lme(thick ~ soil, data=spdata, correlation = corGaus(form = ~ east + north), random= ~1|dummy, method="ML")
summary(modelGausFinal)
```
Now in the model with Gaussian spatial structure variable soil isn't statistically significant as p-value is big.

## Conclusions
Without including the spatial structure in the model, quality of soil is a statistically significant predictor of thickness. After adding the spatial structure, this relationship becomes not significant. 
