---
title: "Homework 12"
author: "Ma³gorzata Pujszo"
date: "Thursday, January 07, 2016"
output: html_document
---

We would like to find out if there is any relation between thich and soil variables in the dataset spdata.

Firstly, let's look at our dataset.

```{r, message=FALSE, warning=FALSE}
library(nlme)
library(geoR)
library(ggplot2)
library(knitr)
spdata <- read.table("http://www.ats.ucla.edu/stat/r/faq/thick.csv", header = T, sep = ",")
head(spdata)
```

I began with a simple model just taking into account those two interesting variables.

```{r}
model1 <- lme(thick ~ soil, data=spdata, random= ~1|dummy, method="ML")
summary(model1)
```

Since the p-value for soil is low, it suggest that there is a relation between those variables. I plotted them together with linear fit.

```{r}
ggplot(spdata, aes(x=soil, y=thick)) + geom_point()+stat_smooth(method="lm")
```

In the plot I don't see clear relation. Since in the dataset we have also information about location of measurement (east and north variables), it may be a good idea to take them into account. 

Let's first look at variograms. (First variogram is showing all information, second is cut in order to show the line.) 

```{r}
plot(Variogram(spdata$thick, dist(spdata[,2:3])),
  xlab="Distance", main="Semivariogram: thick")
plot(Variogram(spdata$thick, dist(spdata[,2:3])),
  xlab="Distance", ylim=c(0, 5), main="Semivariogram: thick")
```

Variograms suggest that distance between measurement may be important. We can see that measurements which were taken closely to each other are the most similar and around distance 60-70 measurements' differences are the largest. 

I would like to take into account spatial structure but I need to find out which spatial correlation structure I should use. In order to check it, let's firstly plot variogram together with different fits.

```{r}
v1 <- variog(coords = spdata[,2:3], data = spdata$thick, message=FALSE)
vexp <- variofit(cov.model="exponential", vario=v1, ini.cov.pars=c(10, 10), message=FALSE )
vgaus <- variofit(cov.model="gaussian", vario=v1, ini.cov.pars=c(10, 10), message=FALSE )
vlin <- variofit(cov.model="linear", vario=v1, ini.cov.pars=c(10, 10), message=FALSE )
vsph <- variofit(cov.model="spherical", vario=v1, ini.cov.pars=c(10, 10), message=FALSE )

par(mar=c(4, 4, 2, 2))
plot(v1, type = "b", main="Variogram") 
lines.variomodel(vexp, col="green")
lines.variomodel(vgaus, col="red")
lines.variomodel(vlin, col="blue")
lines.variomodel(vsph, col="orange")
legend(90,2.5, c("Exponential", "Gaussian", "Linear", "Spherical"), lty=c(1,1,1,1,1), col=c("green", "red", "blue", "orange"), cex=0.7)
```

The best fit seem to have Gaussian structure. But let's build those models and see which fits best (considering AIC, BIC and LogLik criteria).

```{r}
aic.models <- vector(length=5)
bic.models <- vector(length=5)
loglik.models <- vector(length=5)
names <- vector(length=5)

model2 <- lme(thick ~ soil, data=spdata, correlation = corGaus(form = ~ east + north), random= ~1|dummy, method="ML")
model3 <- lme(thick ~ soil, data=spdata, correlation = corSpher(form = ~ east + north), random= ~1|dummy, method="ML")
model4 <- lme(thick ~ soil, data=spdata, correlation = corLin(form = ~ east + north), random= ~1|dummy, method="ML")
model5 <- lme(thick ~ soil, data=spdata, correlation = corExp(form = ~ east + north), random= ~1|dummy, method="ML")

models <- list(model1, model2, model3, model4, model5)

for (i in 1:5) {
  model <- models[[i]]
  aic.models[i] <- summary(model)$AIC
  bic.models[i] <- summary(model)$BIC
  loglik.models[i] <- summary(model)$logLik 
  names[i] <- paste0("model", i)
}
kable(cbind(names, aic.models))
kable(cbind(names, bic.models))
kable(cbind(names, loglik.models))
```

All three criteria gives the same answer that model2 (with Gaussian correlation structure is the best fit). Consequently, I used it as a final model.

```{r}
summary(model2)
```

Large p-value of soil variable indicates that there is no relation between thich and soil. This example shows that without taking into account spatial structure of measurements we would wrongly conclude that there is a relation between interesting variables whereas in fact there isn't any.