---
title: "Praca domowa 12"
author: "Piotr Obarski"
date: "Modele liniowe i mieszane"
output: 
  html_document:
  toc: TRUE
---

##Goal
Try two models: simple linear regression model with just one dependent variable and a model which takes into account spatial structure of measurements (variables east and north).


Compare these two models and answer the question: is there a significant relation between variables thick (y) and soil (dependent variable)?
##Solution

Let's load the data and see the first few rows. I also plot the thick and soil columns.

```{r}
library(nlme)

spdata <- read.table("http://www.ats.ucla.edu/stat/r/faq/thick.csv", header = T, sep = ",")
head(spdata)
plot(thick~soil, data=spdata)
```

We see that there can be a linear relationship between these two variables.
Firstly, build the model with fixed effects.

```{r}
model0 <- lme(fixed = thick ~ soil, data = spdata, random = ~ 1 | dummy, method = "ML")
summary(model0)
```

We see that soil is important variable.

Now let's build the second model, which takes into account spatial structure.

```{r}
model1 <- lme(fixed = thick ~ soil, data=spdata, correlation = corGaus(form = ~ east + north), random= ~1|dummy, method="ML")
summary(model1)

```

Here we see that p-value for soil variable is very high, so it is not important (i.e. information about soil is mostly included in spatial structure).

##Conclusions

Looking at measures of model fit (AIC, BIC, logLik) we see that the second model is better. Also in the first model soil is significant while in the second is not.