---
title: "Kinship effects"
author: "Przemyslaw Biecek"
date: "Linear models with fixed and random effects"
output: 
  html_document:
    toc: TRUE
---

# Mice again

Let's find out what has an impact on maximal stress of muscle in Mus Musculus (mice).

```{r, message=FALSE, warning=FALSE}
library(lme4)
library(PBImisc)
library(ggplot2)
library(coxme)
require(kinship2)

head(musculus)
```

# So simple

```{r}
ggplot(musculus, aes(diet, sigma)) + 
  geom_boxplot() + geom_jitter(color="grey") + theme_bw()
  
summary(lm(sigma~diet, data=musculus))
```

# Yet observations are not independent

The kinship coefficient is a measure of relatedness between two individuals. It represents the probability that two alleles, sampled at random from each individual are identical.

```{r}
gped <- with(musculus, pedigree(id, dadid, momid, sex=sex))
plot(gped)

kmat <- kinship(gped)
kmat[1:5,1:15]
```

# Let's add polygenetic effect to the model

```{r}
mfit <- lmekin(sigma~diet + (1|id), data=musculus, varlist=kmat*2)
mfit

# Let's compare with
summary(lm(sigma~diet, data=musculus))
```

Percentage of inherited variance.

Let's extract variances and then we can see how large is the shared variance.

```{r}
(sigmaa <- mfit$vcoef$id)
(sigma0 <- mfit$sigma^2)

sigmaa/(sigmaa+sigma0)
```

# Other applications

Meta analysis, where data is pulled from different studies.

```{r}
head(dementia)
```

We allow fosome individual effects in each study, but overall we are interested in the main effect shared between different studies.

```{r}
modelFull <- lmer(demscore~age+sex+(age+sex|study), data=dementia, REML=F)
```

Fixed effects and random deviations

```{r}
modelFull
random.effects(modelFull)
```

Different physicians in different centers

```{r}
head(eden[,c(1,2,7,8,9,12)])

library(lattice)
bwplot(BPRS.Average~mdid|center, data=eden, scales=list(x=list( relation="free")))

summary(model <- lm(MANSA~day, data=eden))
summary(model <- lmer(MANSA~day + (1|center/mdid), data=eden, REML=FALSE))
```


# That's almost the end

What you should remember in modelling

* What about transformations for y?
* What about transformations for x?
* Difference between crossed / nested effects
* Difference between fixed / random effects
* How to estimate coefficient?
* How to derive confidence intervals for effects?
* How to calculate predictions (point / interval)?
* Model diagnostic
* How to test effects?

