---
title: "Homework Week 14"
author: "Annanina Koster"
date: "18 januari 2016"
output: html_document
---


---


The homework [21 I 2016]

We will work with data about previous homeworks.

For each homework let's count the number of lines in the Rmd file. Then take into account two other variables: the id (number / week) of the homework and the person that is doing the homework.

Create a model that explains the number of lines of homework. What is the relation between number of lines of the homework and the id / week of the homework?

Should the author/submitter be modelled as a fixed or random effect?


---


First, by looking at all the homeworks handed in on github, I created a dataset saying the number of the homework, the author and the amount of lines in the rmd file. 

```{r}

H1.AK <- c(1, "AK", 81)
H1.EP <- c(1, "EP", 74)
H1.KK <- c(1, "KK", 48)
H1.KR <- c(1, "KR", 52)
H1.MF <- c(1, "MF", 61)
H1.MB <- c(1, "MB", 50)
H1.AS <- c(1, "AS", 54)
H1.KG <- c(1, "KG", 67)
H1.MP <- c(1, "MP", 42)
H1.MK <- c(1, "MK", 47)
H1.MaPa <- c(1, "MaPa", 60)
H1.PO <- c(1, "PO", 75)
H1.PiOb <- c(1, "PiOb", 42)
H1.RC <- c(1, "RC", 39)
H1.MW <- c(1, "MW", 35)

H2.AK <- c(2, "AK", 39)
H2.KK <- c(2, "KK", 177)
H2.AS <- c(2, "AS", 81)
H2.EP <- c(2, "EP", 127)
H2.MP <- c(2, "MP", 82)
H2.KR <- c(2, "KR", 67)
H2.MaPa <- c(2, "MaPa", 50)
H2.PO <- c(2, "PO", 76)
H2.PiOb <- c(2, "PiOb", 111)

H3.AK <- c(3, "AK", 83)
H3.KK <- c(3, "KK", 212)
H3.MB <- c(3, "MB", 113)
H3.MF <- c(3, "MF", 53)
H3.AS <- c(3, "AS", 173)
H3.EP <- c(3, "EP", 133)
H3.KG <- c(3, "KG", 116)
H3.MP <- c(3, "MP", 105)
H3.MaPa <- c(3, "MaPa", 94)
H3.RC <- c(3, "RC", 79)
H3.PiOb <- c(3, "PiOb", 91)

H4.PB <- c(4, "PB", 116)
H4.KK <- c(4, "KK", 284)
H4.MB <- c(4, "MB", 115)
H4.MF <- c(4, "MF", 122)
H4.AS <- c(4, "AS", 139)
H4.AK <- c(4, "AK", 134)
H4.EP <- c(4, "EP", 94)
H4.KG <- c(4, "KG", 141)
H4.KR <- c(4, "KR", 156)
H4.MP <- c(4, "MP", 172)
H4.PiOb <- c(4, "PiOb", 100)
H4.MW <- c(4, "MW", 92)

H5.MF <- c(5, "MF", 55)
H5.KK <- c(5, "KK", 163)
H5.MB <- c(5, "MB", 69)
H5.AS <- c(5, "AS", 61)
H5.AK <- c(5, "AK", 116)
H5.BP <- c(5, "BP", 42)
H5.EP <- c(5, "EP", 111)
H5.KR <- c(5, "KR", 54)
H5.MP <- c(5, "MP", 60)
H5.PiOb <- c(5, "PiOb", 67)
H5.RC <- c(5, "RC", 56)

H6.MF <- c(6, "MF", 73)
H6.AS <- c(6, "AS", 136)
H6.AK <- c(6, "AK", 216)
H6.EP <- c(6, "EP", 123)
H6.KK <- c(6, "KK", 229)
H6.PiOb <- c(6, "PiOb", 146)

H7.AS <- c(7, "AS", 186)
H7.AK <- c(7, "AK", 442)
H7.EP <- c(7, "EP", 183)
H7.KK <- c(7, "KK", 323)
H7.MP <- c(7, "MP", 149)
H7.MW <- c(7, "MW", 122)
H7.PiOb <- c(7, "PiOb", 184)

H8.MB <- c(8, "MB", 64)
H8.AS <- c(8, "AS", 53)
H8.AK <- c(8, "AK", 86)
H8.EP <- c(8, "EP", 71)
H8.KK <- c(8, "KK", 158)
H8.MP <- c(8, "MP", 82)
H8.MW <- c(8, "MW", 52)
H8.PiOb <- c(8, "PiOb", 177)

H9.MB <- c(9, "MB", 78)
H9.AS <- c(9, "AS", 68)
H9.EP <- c(9, "EP", 83)
H9.KK <- c(9, "KK", 136)
H9.MP <- c(9, "MP", 73)
H9.PiOb <- c(9, "PiOb", 42)
H9.AK <- c(9, "AK", 73)

H10.AS <- c(10, "AS", 60)
H10.EP <- c(10, "EP", 59)
H10.KK <- c(10, "KK", 149)
H10.MP <- c(10, "MP", 46)
H10.PiOb <- c(10, "PiOb", 27)
H10.AK <- c(10, "AK", 50)

H11.MF <- c(11, "MF", 79)
H11.PiOb <- c(11, "PiOb", 60)
H11.MB <- c(11, "MB", 87)
H11.AS <- c(11, "AS", 68)
H11.EP <- c(11, "EP", 91)
H11.KK <- c(11, "KK", 159)
H11.MP <- c(11, "MP", 63)
H11.AK <- c(11, "AK", 117)

H12.MF <- c(11, "MF", 89)
H12.KK <- c(11, "KK", 97)
H12.AS <- c(11, "AS", 82)
H12.EP <- c(11, "EP", 108)
H12.MP <- c(11, "MP", 98)
H12.PiOb <- c(11, "PiOb", 49)
H12.AK <- c(11, "AK", 73)
H12.MB <- c(11, "MB", 95)

own.data <- rbind(H1.AK, H1.EP, H1.KK, H1.KR, H1.MF, H1.MB, H1.AS, H1.KG, H1.MP, H1.MK, H1.MaPa, H1.PO, H1.PiOb, H1.RC, H1.MW, H2.AK, H2.KK, H2.AS, H2.EP, H2.MP, H2.KR, H2.MaPa, H2.PO, H2.PiOb, H3.AK, H3.KK, H3.MB, H3.MF, H3.AS, H3.EP, H3.KG, H3.MP, H3.MaPa, H3.RC, H3.PiOb, H4.PB, H4.KK, H4.MB, H4.MF, H4.AS, H4.AK, H4.EP, H4.KG, H4.KR, H4.MP, H4.PiOb, H4.MW, H5.MF, H5.KK, H5.MB, H5.AS, H5.AK, H5.BP, H5.EP, H5.KR, H5.MP, H5.PiOb, H5.RC, H6.MF, H6.AS, H6.AK, H6.EP, H6.KK, H6.PiOb, H7.AS, H7.AK, H7.EP, H7.KK, H7.MP, H7.MW, H7.PiOb, H8.MB, H8.AS, H8.AK, H8.EP, H8.KK, H8.MP, H8.MW, H8.PiOb, H9.MB, H9.AS, H9.EP, H9.KK, H9.MP, H9.PiOb, H9.AK, H10.AS, H10.EP, H10.KK, H10.MP, H10.PiOb, H10.AK, H11.MF, H11.PiOb, H11.MB, H11.AS, H11.EP, H11.KK, H11.MP, H11.AK, H12.MF, H12.KK, H12.AS, H12.EP, H12.MP, H12.PiOb, H12.AK, H12.MB)
own.data1 <- as.data.frame(own.data)

```

In order to estimate a model for these data, it should be noticed that for each week a subset of all the authors handed in a homework assignment, but the homeworks are different every time. Hence, the author should be nested within the week of the homework. Hence, author should be modelled as a random effect. Moreover, we are interested in the relation between the number of lines (V3) and the number of the homework (V1). Therefore, V1 should be modeled as a fixed effect. No correlation should be present between the variables, so no structure correcting for this has to be used. The V1 variable does not need a log transfer because it is very nicely ordered already. As can be seen in the histograms, also the V3 variable does not really need a log-transform. There is not much to visualize in this dataset. But the plot shows that there is no strong pattern between the number of the homework 

```{r}

library(lattice)
xyplot(V3 ~ V1 | V2, data=own.data1, type = c("g","p","r"), index = function(x,y) coef(lm(y ~ x))[1])

hist(own.data1$V3)
hist(log(own.data1$V3))

library(lme4)
own.data1$V1 <- as.numeric(own.data1$V1)
own.data1$V2 <- as.factor(own.data1$V2)
own.data1$V3 <- as.numeric(own.data1$V3)
model1 <- lmer(V3 ~ V1 + (1|V1/V2), data=own.data1)
summary(model1)
AIC(model1)
BIC(model1)
logLik(model1)
model2 <- lm(V3 ~ V1/V2, data=own.data1)
summary(model2)
AIC(model2)
BIC(model2)
logLik(model2)
model3 <- lmer(V3 ~ V1 + (V1|V2), data=own.data1)
summary(model3)
AIC(model3)
BIC(model3)
logLik(model3)

model3u = ranef(model1, condVar =TRUE)
dotplot(u)

```

As you can see, the variation per homework (week) is quite high. The second estimated model is without a random effect. It just estimates the effect of the week of the homework per person. Except for the intercept, none of the variables is significant. The AIC, BIC and logLik also show that model1 is the better model. Also when compared to model3. Since the prediction plot seems to follow a normal distribution, it can be concluded that this is a good model. 

In conclusion, the effect of the week of the homework on the number of lines is slightly negative. The author has been modeled as a random effect.




